---
title: "Data Mining Project 1"
author: "Project Group 10 - Abhilash Narayanan"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

```{r}
library("tidyverse")
library("ggplot2")
library("ggrepel")
library("ggcorrplot")
library("DT")
```
# 1. Data Collection and Cleaning

Three datasets are considered for this analysis

1. Dataset 1 : COVID-19_cases_TX.csv
2. Dataset 2 : COVID-19_cases_plus_census.csv
    ( Cleansing of this is done in a R code uploaded separately)
3. Dataset 4 : COVID-19_Vaccinations  

```{r}
tx_covid_cases <-
read.csv("C:\\Users\\nilee\\Documents\\AbhilashStudy\\DataMining\\COVID-19\\COVID-19_cases_TX.csv",header =
TRUE)
options(max.print=10)

county_census <-
read.csv("C:\\Users\\nilee\\Documents\\AbhilashStudy\\DataMining\\COVID-19\\COVID-19_cases_plus_census.csv",header =
TRUE)
options(max.print=10)

county_vaccination <-
read.csv("C:\\Users\\nilee\\Documents\\AbhilashStudy\\DataMining\\COVID-19\\COVID-19_Vaccinations_in_the_United_States_County.csv",header =
TRUE)
options(max.print=10)

```

Changing the options to print only 10 records while printing
Printing summary of COVID dataset

```{r}
options(max.print=10)
summary(tx_covid_cases)
```
Printing the Summary of the Vaccination Dataset

```{r}
summary(county_vaccination)
```

## Data Cleaning Dataset 1

Below are the cleaning activities performed for this dataset:-
Removing the records which are Un-allocated to a county as these may not have a value while performing a county level analysis
Correcting the data types of the data columns

```{r}

tx_covid_cases

tx_covid_cases <- subset(tx_covid_cases,county_name !='Statewide Unallocated') 
tx_covid_cases$date <- as.Date(tx_covid_cases$date,format = "%m/%d/%Y")
options(max.print=10)
tx_covid_cases
```

## Data Cleaning Dataset 4

Correcting the data type of the "Date" feature
Changing the data type of all numeric fields by removing commas

```{r}
county_vaccination$Date <- as.Date(county_vaccination$Date,format = "%m/%d/%Y")
county_vaccination$Administered_Dose1_Recip <- as.numeric(gsub(",","",county_vaccination$Administered_Dose1_Recip,fixed=TRUE))
county_vaccination$Series_Complete_Yes <- as.numeric(gsub(",","",county_vaccination$Series_Complete_Yes,fixed=TRUE))
county_vaccination$Booster_Doses <- as.numeric(gsub(",","",county_vaccination$Booster_Doses,fixed=TRUE))
county_vaccination$Census2019 <- as.numeric(gsub(",","",county_vaccination$Census2019,fixed=TRUE))
options(max.print=10)
county_vaccination
```


Summary of the COVID cases and Death by County

```{r}
tx_covid_cases_county = subset(tx_covid_cases, date == '2021-01-25') %>% group_by(county_name)  %>%
                    summarise(total_confirmed_cases = sum(confirmed_cases), 
                              total_deaths = sum(deaths), 
                              .groups = 'drop')
tx_covid_cases_county$death_perc=tx_covid_cases_county$total_deaths*100/tx_covid_cases_county$total_confirmed_cases 
tx_covid_cases_county <- tx_covid_cases_county[order(tx_covid_cases_county$death_perc,
                                decreasing = TRUE), ]
tx_covid_cases_county

```


# 2. Data Analysis

## CoVID Data Set

```{r}
ggplot(head(tx_covid_cases_county,10),
           aes(x=county_name, y=death_perc,fill=county_name)) + 
  geom_bar(stat="identity",width=0.5) +
  ggtitle("Counties in TX \n Highest Death Percentage \n w.r.t confirmed cases")+
  coord_flip()
head

```
```{r}
ggplot(tail(tx_covid_cases_county,10),
           aes(x=county_name, y=death_perc,fill=county_name)) + 
  geom_bar(stat="identity",width=0.5) +
  ggtitle(" 10 Counties in TX \n (Lowest Death Percentage w.r.t confirmed cases)")+
  coord_flip()
head
```
Creating the County dataset from Map for Tx state
This will be joined with the Cases dataset to plot the same in the graph

```{r}
counties <- as_tibble(map_data("county"))
counties_TX <- counties %>% dplyr::filter(region == "texas") %>% rename(c(county = subregion))

tx_covid_cases_county <- tx_covid_cases_county %>% mutate(county = county_name %>% str_to_lower() %>% 
    str_replace('\\s+county\\s*$', ''))

counties_w_map<-merge(x = counties_TX, y = tx_covid_cases_county, by = "county")


ggplot(counties_w_map, aes(long, lat, label = county)) + 
  geom_polygon(aes(group = group, fill = death_perc)) +
  scale_fill_gradient(low="green", high="red") +
  coord_quickmap() +
  labs(title = "Death Percentage per County")
```


### Filtering only TX Counties for Census data

```{r}
county_census <- subset(county_census,state == 'TX')
county_census <- county_census %>% mutate(county = county_name %>% str_to_lower() %>% 
    str_replace('\\s+county\\s*$', ''))
```

### Joining the Covid data with the Census data

```{r}
covid_cases_w_census<-merge(x = county_census, y = tx_covid_cases_county, by = "county")
```

### Is there a relation between median income and death percentage of a county?

```{r}
ggplot(covid_cases_w_census, mapping = aes(x= median_income, y = death_perc, label = county_name.x)) + 
  geom_smooth(method = lm) +
  geom_point(mapping = aes(size = total_confirmed_cases), color = "blue") + 
  geom_text_repel(data = subset(covid_cases_w_census, death_perc > quantile(death_perc, .95))) +
  labs(x = "Median Income", y = "Death Percentage", size = "Confirmed Cases")
```
```{r}
ggplot(covid_cases_w_census, mapping = aes(x= total_pop, y = median_income, label = county_name.x)) + 
  geom_smooth(method = lm) +
  geom_point(mapping = aes(size = death_perc), color = "violet") + 
  geom_text_repel(data = subset(covid_cases_w_census, death_perc > quantile(death_perc, .95))) +
  labs(x = "Median Income", y = "Total Population", size = "Death Percentage")
```
Total Population do not have much effect on teh death percentage compared to the median income. 
If the MEdian Income of a county is lower , irrespective of the total population , these counties seem to have a higher death percentage


### Did we flatten the curve? 

Logic to add total confirmed cases - previous day total confirmed cases

```{r}

tx_covid_cases_date = tx_covid_cases %>% group_by(date) %>%
summarise(total_confirmed_cases = sum(confirmed_cases),
total_deaths = sum(deaths),
.groups = 'drop')
tx_covid_cases_date <- mutate(tx_covid_cases_date, prev_day_case=lag(total_confirmed_cases, order_by = date))
tx_covid_cases_date$daily_increase<- tx_covid_cases_date$total_confirmed_cases-tx_covid_cases_date$prev_day_case
tx_covid_cases_date

```

Plotting that in a line graph

```{r}
ggplot(tx_covid_cases_date, aes(x = date, y = daily_increase)) + 
  geom_line() + 
  geom_smooth() +
  labs(y = "New Cases per Day")
```
  
### Finding Corelated features  

```{r}
cor_cases=cor(select(covid_cases_w_census,confirmed_cases,total_pop,percent_income_spent_on_rent, confirmed_cases,commute_45_59_mins,income_less_10000,commuters_by_public_transportation,income_per_capita,death_perc))
ggcorrplot(cor_cases,insig = "blank", hc.order = TRUE)
```
## Dataset 4: COVID Vaccination Report

Pivoting the data to display multiple line graphs

```{r}

county_vacc_date = county_vaccination %>% group_by(Date) %>%
summarise(tot_dose1 = sum(Administered_Dose1_Recip),
        tot_series_complete = sum(Series_Complete_Yes),
        tot_booster = sum(Booster_Doses),
.groups = 'drop')

county_vacc_date_pivot <- county_vacc_date %>%
  select(Date, tot_dose1, tot_series_complete,tot_booster) %>%
  gather(key = "vaccination_type", value = "value", -Date)
county_vacc_date_pivot
```

Plotting the pivoted data set to a line graph

```{r}

ggplot(county_vacc_date_pivot, aes(x = Date, y = value)) + 
  geom_line(aes(color = vaccination_type, line_type = 1)) + 
  scale_color_manual(values = c("darkred", "darkgreen","blue")) +
  labs(y = "CoVID Vaccinations", x="Date", title="COVID Vaccination Counts")+
  coord_flip()
  

```
### Calculating the Booster Doses /1000

```{r}
tx_covid_vacc_county <- county_vaccination %>% mutate(county = Recip_County %>% str_to_lower() %>% 
    str_replace('\\s+county\\s*$', ''))

tx_covid_vacc_county = subset(tx_covid_vacc_county, Date == '2023-05-10') 

tx_covid_vacc_county$Booster_per_1000 = (tx_covid_vacc_county$Booster_Doses*1000)/ tx_covid_vacc_county$Census2019

vacc_w_map<-merge(x = counties_TX, y = tx_covid_vacc_county, by = "county")

vacc_w_map$Booster_per_1000 = vacc_w_map$Booster_Doses*1000/vacc_w_map$Census2019

vacc_w_map

```

### Plotting the Booster_per_1000 doses on a map

```{r}
ggplot(vacc_w_map, aes(long, lat, label = county)) + 
  geom_polygon(aes(group = group, fill = Booster_per_1000)) +
  scale_fill_gradient(low="lightgreen", high="darkgreen") +
  coord_quickmap() +
  labs(title = "Booster Doses per 1000")
```
### Merging this with the Census data

```{r}
covid_vacc_w_census<-merge(x = county_census, y = tx_covid_vacc_county, by = "county")
```

### Is there a relation for Booster vaccines and Median Income?

```{r}
ggplot(covid_vacc_w_census, mapping = aes(x= income_per_capita, y = Booster_per_1000, label = county)) + 
  geom_smooth(method = lm) +
  geom_point(mapping = aes(size = total_pop), color = "lightgreen") + 
  geom_text_repel(data = subset(covid_vacc_w_census, Booster_per_1000 > quantile(Booster_per_1000, .95))) +
  labs(x = "Income per capita", y = "Booster_per_1000", size = "Total Population")
```
It seems like the Booster vaccine shots do not have 
direct co-relation between Income and total population 


```{r}
covid_vacc_w_cases<-merge(x = tx_covid_cases_county, y = tx_covid_vacc_county, by = "county")
```

### Can we establish any relationship with Booster shots and COVID cases?

```{r}
ggplot(covid_vacc_w_cases, mapping = aes(x= Booster_per_1000, y = death_perc, label = county)) + 
  geom_smooth(method = lm) +
  geom_point(mapping = aes(size = total_deaths), color = "lightgreen") + 
  geom_text_repel(data = subset(covid_vacc_w_cases, death_perc > quantile(death_perc, .95))) +
  labs(x = "Booster_per_1000", y = "Death Percentage", size = "Total Population")
```



